---
title: "Species Richness Estimation Report"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
params:
  taxon_name: "Unnamed taxon"
  mode: "parasitic"
  interval_years: 5
  results: !r NULL
  interval_table: !r NULL
  generated_at: !r Sys.time()
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, fig.width = 8, fig.height = 5)
# plot_*() functions are already available in .GlobalEnv (sourced once at app
# startup by global.R); no need to re-source here, and the report's own
# working directory is report/, not the project root, so a relative
# source("R/plotting.R") would fail anyway.
results <- params$results
interval_table <- params$interval_table
```

# `r params$taxon_name`

**Mode:** `r ifelse(params$mode == "parasitic", "Parasitic (host-availability covariate available)", "Free-living (no host covariate)")`
**Interval width:** `r params$interval_years` years
**Report generated:** `r format(params$generated_at, "%Y-%m-%d %H:%M %Z")`

*This report was generated on demand from data you supplied for this session only. No
uploaded data, fitted model, or figure is stored on the server; this document is the
only durable record of the analysis.*

## Method

New species described in interval $i$ ($S_i$) are modeled as
$S_i \sim \text{Poisson}(\lambda_i)$, with

$$\log \lambda_i = \big(\log L_0 + \beta \cdot Y_i\big) + \log(S_T - \text{cumS}_i) \; [+ \log(H_t - \text{cumH}_i)] \;+ \log(1 + T_i)$$

where $S_T$ is the total number of species expected to exist, $L_0$ and $\beta$
describe baseline and time-trending taxonomic efficiency, $T_i$ is the number of
active taxonomists (an effort proxy), $\text{cumS}_i$ is the cumulative species
described before interval $i$, and the optional host term $(H_t - \text{cumH}_i)$
is included only for the "host" structure. This is the same model family used in
Boeger et al. (submitted, *Zoologia*, ZOOL-2026-0012.R1) to estimate Brazilian
Dactylogyridae and Actinopterygii richness, generalized here to any user-supplied
taxon and adapted from Pimm et al. (2010) and Joppa et al. (2011). Models were fit
in Stan via `rstan`, with priors on $S_T$ (Gamma), $\log L_0$ and $\beta$ (Normal)
supplied by the user for this run.

## Data summary

```{r}
knitr::kable(interval_table, digits = 2)
```

```{r}
knitr::kable(
  data.frame(
    Metric = c("Total intervals", "Total species described (observed)", "Total unique taxonomists (summed)"),
    Value = c(nrow(interval_table), sum(interval_table$n_species), sum(interval_table$n_authors))
  )
)
```

```{r effort-scatter}
plot_effort_scatter(interval_table)
```

```{r single-results, results='asis', eval=identical(results$job_type, "single")}
structure_labels <- c(host = "With host covariate", no_host = "Without host covariate")
for (st in names(results$per_structure)) {
  r <- results$per_structure[[st]]
  cat("\n\n## Model:", structure_labels[[st]], "\n\n")

  cat("\n\n### Parameter estimates\n\n")
  print(knitr::kable(r$summary, digits = 4, row.names = FALSE))

  cat("\n\n### Posterior distributions\n\n")
  print(plot_posterior_hist(r$fit, "ST"))
  print(plot_posterior_hist(r$fit, "log_L0"))
  print(plot_posterior_hist(r$fit, "beta"))

  cat("\n\n### Trace plots (convergence diagnostics)\n\n")
  print(plot_trace(r$fit, "ST"))

  cat("\n\n### Simulated vs. observed discovery rate\n\n")
  print(plot_ppc(r$fit, interval_table))

  cat("\n\n### Percent of estimated total diversity already described\n\n")
  print(plot_percent_described(r$pct_draws))

  cat("\n\n### Extrapolated cumulative discovery\n\n")
  st_median <- r$summary$median[r$summary$parameter == "ST"]
  print(plot_extrapolation(interval_table, r$extrapolation, ST_median = st_median))

  cat("\n\n")
}
```

```{r battery-results, results='asis', eval=identical(results$job_type, "battery")}
cat("\n\n## Sensitivity battery: all scenarios \u00d7 structures\n\n")
cat("\n\nBest-supported model by LOO-ELPD: **", results$best_model_label, "**\n\n")

comp <- results$comparison_table
print(knitr::kable(comp[, c("model_label", "scenario", "structure", "ST_median", "ST_lower", "ST_upper",
                             "elpd_loo", "elpd_diff", "mean_crps", "mean_mae")], digits = 3, row.names = FALSE))

print(plot_loo_comparison(comp))
print(plot_predictive_scores(comp))

cat("\n\n### Detail for the best-supported model\n\n")
best_fit <- results$fits[[results$best_model_label]]
print(plot_posterior_hist(best_fit, "ST"))
print(plot_ppc(best_fit, interval_table))
```

## Data handling notice

No dataset, credential, or result from this analysis is retained by the
application after this browser session ends. Save this report and any
downloaded figures locally if you need them again.
