---
title: "Species Richness Estimation Report"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
params:
  taxon_name: "Unnamed taxon"
  mode: "parasitic"
  interval_years: 5
  results: !r NULL
  interval_table: !r NULL
  generated_at: !r Sys.time()
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, fig.width = 8, fig.height = 5)
# plot_*()/structure_label() functions are already available in .GlobalEnv
# (sourced once at app startup by global.R); no need to re-source here, and
# the report's own working directory is report/, not the project root, so a
# relative source("R/plotting.R") would fail anyway.
results <- params$results
interval_table <- params$interval_table
```

# `r params$taxon_name`

**Mode:** `r ifelse(params$mode == "parasitic", "Parasitic (host structures available)", "Free-living (no-host structures only)")`
**Interval width:** `r params$interval_years` years
**Report generated:** `r format(params$generated_at, "%Y-%m-%d %H:%M %Z")`

*This report was generated on demand from data you supplied for this session only. No
uploaded data, fitted model, or figure is stored on the server; this document is the
only durable record of the analysis.*

## Method

New species described in interval $i$ ($S_i$) are modeled as
$S_i \sim \text{Poisson}(\lambda_i)$, with

$$\log \lambda_i = \log L_i + \log(S_T - \text{cumS}_i) \; [+ \log(H_t - \text{cumH}_i)] \;+ \log T_i$$

where $S_T$ is the total number of species expected to exist, $T_i$ is the
number of active taxonomists (an effort proxy), $\text{cumS}_i$ is the
cumulative species described before interval $i$, and the optional host term
$(H_t - \text{cumH}_i)$ is included only for "host" structures. The
efficiency term $L_i$ takes one of two forms depending on the structure's
trend form: **exponential**, $L_i = L_0 \cdot e^{\beta Y_i}$ (fit as
$\log L_i = \log L_0 + \beta Y_i$), or **linear**, $L_i = L_0 + \beta Y_i$
(floored at a small positive value). This is the same model family used in
Boeger et al. (submitted, *Zoologia*, ZOOL-2026-0012.R1) to estimate Brazilian
Dactylogyridae (host + exponential) and Actinopterygii (no-host + linear)
richness, generalized here to any user-supplied taxon and adapted from Pimm
et al. (2010) and Joppa et al. (2011). Models were fit in Stan via `rstan`,
with priors on $S_T$ (Gamma) and $\beta$ (Normal), plus $\log L_0$ or $L_0$
(Normal, whichever the fitted structure's trend form needs), supplied by the
user for this run. *Implementation note:* to keep $\log T_i$ finite for any
interval with zero active taxonomists, the fitted model evaluates
$\log(1+T_i)$ rather than $\log T_i$, a standard numerical safeguard that is
negligible whenever $T_i \geq$ a few.

## Data summary

```{r}
knitr::kable(interval_table, digits = 2)
```

```{r}
knitr::kable(
  data.frame(
    Metric = c("Total intervals", "Total species described (observed)", "Total unique taxonomists (summed)"),
    Value = c(nrow(interval_table), sum(interval_table$n_species), sum(interval_table$n_authors))
  )
)
```

```{r effort-scatter}
plot_effort_scatter(interval_table)
```

```{r single-comparison, results='asis', eval=identical(results$job_type, "single") && !is.null(results$comparison_table)}
cat("\n\n## Structure comparison\n\n")
cat("\n\nYou fit", length(results$per_structure), "structures on the same data \u2014 ranked below by LOO-ELPD (higher/less negative is better). This is how the app lets the observed S_i-vs-effort-vs-time relationship indicate which formula (linear or exponential trend, with or without the host term) fits best, rather than requiring a guess.\n\n")
cat("\n\nBest-supported: **", structure_label(results$best_model_label), "**\n\n")

comp <- results$comparison_table
print(knitr::kable(comp[, c("model_label", "structure", "ST_median", "ST_lower", "ST_upper",
                             "elpd_loo", "elpd_diff", "mean_crps", "mean_mae")], digits = 3, row.names = FALSE))
print(plot_loo_comparison(comp))
print(plot_predictive_scores(comp))
```

```{r single-results, results='asis', eval=identical(results$job_type, "single")}
for (st in names(results$per_structure)) {
  r <- results$per_structure[[st]]
  cat("\n\n## Model:", structure_label(st), "\n\n")

  cat("\n\n### Parameter estimates\n\n")
  print(knitr::kable(r$summary, digits = 4, row.names = FALSE))

  cat("\n\n### Posterior distribution and trace plot (S_T)\n\n")
  print(plot_posterior_hist(r$fit, "ST"))
  print(plot_trace(r$fit, "ST"))

  cat("\n\n### Simulated vs. observed discovery rate\n\n")
  print(plot_ppc(r$fit, interval_table))

  cat("\n\n### Percent of estimated total diversity already described\n\n")
  print(plot_percent_described(r$pct_draws))

  cat("\n\n### Extrapolated cumulative discovery\n\n")
  st_median <- r$summary$median[r$summary$parameter == "ST"]
  print(plot_extrapolation(interval_table, r$extrapolation, ST_median = st_median))

  cat("\n\n")
}
```

```{r battery-results, results='asis', eval=identical(results$job_type, "battery")}
cat("\n\n## Sensitivity battery: all scenarios \u00d7 structures\n\n")
cat("\n\nBest-supported model by LOO-ELPD: **", structure_label(sub("^[^_]+_", "", results$best_model_label)) , "** (scenario ", sub("_.*$", "", results$best_model_label), ")\n\n")

comp <- results$comparison_table
print(knitr::kable(comp[, c("model_label", "scenario", "structure", "ST_median", "ST_lower", "ST_upper",
                             "elpd_loo", "elpd_diff", "mean_crps", "mean_mae")], digits = 3, row.names = FALSE))

print(plot_loo_comparison(comp))
print(plot_predictive_scores(comp))

cat("\n\n### Detail for the best-supported model\n\n")
best_fit <- results$fits[[results$best_model_label]]
print(plot_posterior_hist(best_fit, "ST"))
print(plot_ppc(best_fit, interval_table))
```

## Data handling notice

No dataset, credential, or result from this analysis is retained by the
application after this browser session ends. Save this report and any
downloaded figures locally if you need them again.
